Behind Bars and Beyond: Investigating Socioeconomic Drivers of Incarceration Rates in the United States

### 2024-04-19 paper/paper.qmd
Made final tweaks to paper.
Made final edits and tweaks to the paper.

Incarceration refers to the state of being confined in prison as punishment for criminal offences [@citeIncarcerationRate], and in examining the United States' prison dilemma, it becomes clear that the crux of the problem is not easily reducible to simple explanations. Instead, it is a complex issue influenced by a multitude of factors including cultural dynamics and socioeconomic conditions. Thus, in an attempt to unravel the intricacies of these factors, this paper seeks to conduct a comparative analysis of global trends in education, unemployment, crime, and poverty rates as independent factors influencing a nation's incarceration. However, recognizing that the total number of incarcerations is heavily dependent on population size, this paper focuses on nations' incarceration rates as the estimand, defined to be the number of incarcerated individuals per 100,000 in a given population [@citeIncarcerationRate], which serves as a better metric for measuring the extent of imprisonment within a society. By exploring the impact of these factors on incarceration rates, we aim to better understand the primary forces behind the United States' high incarceration rate and overall number of individuals incarcerated. This understanding will help inform policy-making decisions aimed at addressing the issue effectively.

@@

The remainder of this paper is structured as follows. @sec-data discusses the raw data, cleaning process, variables of interest, and offers visual representations of the data through tabular and graphical representations of the observations. @sec-model introduces and justifies the Bayesian linear regression model used in the analysis of the data in predicting the relationship between the aforementioned independent variables and a nation's incarceration rate. @sec-results deals with analyzing the trends and correlations showcased by the model in more detail. @sec-discussion discusses the real-world implications of the results uncovered in the prior sections, and finally @sec-limitations discusses the limitations and weaknesses of the analysis conducted, as well as the next steps that could be taken to improve the overall reliability of the paper.

@@

The finalized cleaned dataset used in the analysis that will follow was created by merging the aforementioned 5 datasets into one. In particular, only countries that were present in all 5 datasets are present in the cleaned dataset, bringing the total number of nations included in this study to 134. Further, the cleaned analysis dataset was filtered to only contain the following information from the 5 datasets: a country's name, its incarceration rate (per 100k population), its national poverty rate, its unemployment rate, its violent crime rate, and its educational ranking category. Note that a nation's violent crime rate and educational ranking category were not explicit data that were provided as part of the 5 raw datasets. Instead, these were manually constructed and appended to the final dataset during the cleaning process.

In this study, a nation's violent crime rate is defined to be the sum of all the crime rates (per 100k population) available through the violent crime dataset in order to have one final value that represents the extent of violent crime in a nation. It is important to note, however, that various nations had missing values for some of the violent crime statistics, and as a result, these values were simply treated as 0 for the sake of the summation. As for the educational ranking category, as opposed to treating educational ranking as a continuous variable in the analysis that will follow, we instead decided to create 10 categories representing the 10 deciles of educational ranks present in the final dataset, based on the ranks of the 134 countries present, essentially splitting the countries into 10 equally sized categories by educational ranking. This was done in order to have a categorical variable whose coefficient will be more easily interpreted in the model present in @sec-model, as interpreting a 1 unit increase in educational ranking can be quite arbitrary, as moving up or down one rank educationally may be more subjective. @tbl-cleaned-data showcases the top 5 countries by incarceration rate in the cleaned dataset.

@@

As it stands, by adhering to the figures above, there does not appear to be a strong correlation between any one of the variables and incarceration rates. In particular, poverty rates and unemployment rates do not showcase any particular trend in any direction. Violent crime on the other hand, does appear to show a somewhat positive correlation with incarceration rates. However what is interesting to note is that the 5 highlighted nations spanning the entire range of incarceration rates do not have a significant range of violent crime rates, at least relative to the overall range of violent crime rates in the entire dataset. The educational ranking category appears to have a negative relationship with incarceration rates, where a higher rank, i.e. a better education, seems to correspond with a lower average incarceration rate. However, somewhat interesting to note is the drop in mean incarceration rates for categories past the peak in the "111-125" ranking category. As such, given the fairly minimal information provided by the graphical and tabular figures above regarding the relationship between the variables of interest and a nation's incarceration rate, we employ a model in @sec-model to come to a more conclusive result.

@@

As the purpose of this paper is the investigation of various socioeconomic factors and their impact on a nation's incarceration rate as a means of better understanding the United States' mass incarceration problem, we will employ a model that will better enable us to infer the relationship between these various factors and incarceration rates. Then using this information, I will attempt to predict what the United States' incarceration rate should be given its own individual data for each of the determinants of incarceration rate, and subsequently uncover whether they are a good estimator of incarceration rate, or whether the problem stems from a different facet.

@@

The particular model that this paper will utilize is a Bayesian multiple linear regression model of incarceration rate as a function of the independent variables present in our analysis. In particular, the model is as follows:

@@

We can see that, somewhat unexpectedly, increases in the poverty rate negatively correlate with incarceration rates, as do increases in unemployment, assuming all else held constant. On the contrary, as predicted, violent crime rates correlate positively with incarceration rates, all else held constant. As for a nation's educational ranking, it appears that there is no clear relationship between education and incarceration rates as nations in the highest category of education, "1-19", would see a much larger increase in predicted incarceration rates as they enter categories near the median educational ranking, before then seeing smaller increase near the lowest ranked nations.

@@

## Nations with extreme incarceration rates necessitate the need to consider additional factors beyond the model's scope {#sec-culture}

@@

The results in @tbl-incarcerationsummary indicated a negative relationship between poverty rates and incarceration rates, suggesting that nations with a generally poorer population, relative to that particular nation's cost of living, and higher income inequality may find themselves with lower incarceration rates. Though this may seem counter-intuitive at first glance, a potential explanation could be that poorer nations are simply less capable of handling crime. In particular, nations may not have the funding necessary to house criminals or hire a sufficient number of individuals to police the nation. For instance Kenya, who feature a poverty rate of 36.1% well above the median but an incarceration rate of 107 fairly below the median, also happen to have a jail occupancy level of 173.2% as of 2022 [@citeKenyaJail]. This could help explain Kenya's relatively low incarceration rate as being the result of an inability to house more inmates.

@@

The possible explanations for this are twofold: Firstly, as was previously examined in @sec-culture, incarceration rates can be heavily dependent on cultural aspects that were not taken into account via the model. As such, certain nations with exceedingly low or exceedingly high incarceration rates may be disproportionately affecting the mean incarceration rates of their respective educational ranking categories, in turn leading to inaccurate conclusions regarding the impact of education on incarceration rates. To illustrate this point further, El Salvador who feature the highest incarceration rate of 1086, also happen to be in the educational ranking category of "111-125", which may explain why the model predicts that this particular category is likely to raise a nation's incarceration rate by the greatest margin.

The second possible explanation could be that nations with poorer levels of education are likely to also be poorer economically, which, as explored in @sec-povertycorruption, could lead to lower incarceration rates, in turn providing a possible explanation for why nations with the lowest levels of education also happen to feature lower incarceration rates than some nations with better education.

@@

By incorporating all insights derived from @sec-actualvspredicted, @sec-povertycorruption, and @sec-educationincarceration, we arrive to the following conclusions for how the United States can best address their issue of mass incarceration:

@@

There are a number of limitations to be addressed with regards to the analysis conducted and subsequent conclusions drawn. A significant limitation pertaining to this study is the lack of up-to-date data across all nations. As was discuss is @sec-data, all raw datasets contained information on nations that may not be accurate as of 2024. In particular, many nations had data that was collected years, and in some cases decades, ago, in turn potentially rendering much of the analysis inaccurate as significant changes could have risen over the years. Moreover, not all datasets included information on all nations, and in many cases, countries that were included in some datasets featured missing data points for various qualities of differing nations. As a result, the subset of countries that were included in this study may not accurately reflect the overarching global trends as only 134 out of the 193 countries recognized by the UN [@citeUN] were taken into account during the analysis. Thus, the subsequent conclusions drawn may be biased as the remaining countries left in the study could be skewed towards more developed nations as these may be more likely to have data readily available. Moreover, due to the missing data points for countries that were included in the study, values such as the violent crime rate statistic that was calculated for each nation is inaccurate as it only included whatever information was available on that particular nation. As a result, nations which may have reported more crime rates than others would in turn bias the results to inaccurately portray such nations as having relatively more crime than they should when compared with nations that had missing data points for certain crime statistics.

A potentially significant limitation also exists with regards to the model used. In particular, the model does not take into account any interaction terms between the various independent variables when it likely should have, given the complex and intricate dynamics between the many variables at play. Though this was done to simplify the complexity of the model to allow for easier interpretations of the results, the caveat was an overall less realistic, and possibly less accurate, model. Another limitation with the model lies in its omission of cultural variations and associated factors across nations. By not grouping countries with similar cultural values or political and judicial policies, the model fails to account for factors that could greatly influence the derived results. As we have already explored in @sec-culture, varying cultural and societal norms with regards to how nations deal with crime can significantly impact their incarceration rates, a fact not taken into account by the model.

@@

Future analysis in the context of this study can be better improved by incorporating more up-to-date data. In particular, web-scraping and compiling the raw data manually as opposed to using the raw datasets used in this study could net an overall more recent, and complete, set of data points. Moreover, incorporating time series data spanning multiple years across all nations would allow us to account for time and country fixed-effects, in turn providing more conclusive data that takes into account the various cultural and societal disparities across nations, and years. Another improvement that could be made is a better analysis of the model itself in order to fine-tune aspects such as the locations of the various priors, as well as exploring the possibility of adding additional factors and interaction terms to more accurately model incarceration rates.

@@

# Model details

@@

We employ the use a posterior predictive check in @fig-ppcheck to evaluate how well the model fits the data. By examining the figure, it appears as though the model does not accurately fit the data, and thus amendments likely should be made as a result [@citeAlexander].

```{r}
#| label: fig-ppcheck
#| fig-cap: Examining how the model fits, and is affected by, the data through a posterior predictive check
#| echo: false
#| fig-pos: H
#| message: false
#| warning: false

@@

```{r}
#| label: fig-postprior
#| fig-cap: Examining how the model fits, and is affected by, the data through a posterior-prior comparison
#| echo: false
#| warning: false
#| fig-pos: H
#| message: false

>>>>

Incarceration refers to the state of being confined in prison as punishment for criminal offences [@citeIncarcerationRate], and in examining the United States' prison dilemma, it becomes clear that the crux of the problem is not easily reducible to simple explanations. Instead, it is a complex issue influenced by a multitude of factors including cultural dynamics and socioeconomic conditions. Thus, in an attempt to unravel the intricacies of these factors, this paper seeks to conduct a comparative analysis of global trends in education, unemployment, violent crime, and poverty rates as independent factors influencing a nation's incarceration. However, recognizing that the total number of incarcerations is heavily dependent on population size, this paper focuses on nations' incarceration rates as the estimand, defined to be the number of incarcerated individuals per 100,000 in a given population [@citeIncarcerationRate], which serves as a better metric for measuring the extent of imprisonment within a society. By exploring the impact of these factors on incarceration rates, we aim to better understand the primary forces behind the United States' high incarceration rate and overall number of individuals incarcerated. This understanding will help inform policy-making decisions aimed at addressing the issue effectively.

@@

The remainder of this paper is structured as follows. @sec-data discusses the raw data, cleaning process, variables of interest, and offers visual representations of the data through tabular and graphical representations of the observations. @sec-model introduces and justifies the Bayesian linear regression model used in the analysis of the data in predicting the relationship between the aforementioned independent variables and a nation's incarceration rate. @sec-results deals with analyzing the trends and correlations showcased by the model in more detail. @sec-discussion discusses the real-world implications of the results uncovered in the prior sections, and finally @sec-limitations discusses the limitations and weaknesses of the analysis conducted, as well as the next steps that could be taken to improve the overall reliability of the paper. Additional information regarding the model details can be found in the Appendix under @sec-model-details.

@@

The finalized cleaned dataset used in the analysis that will follow was created by merging the aforementioned 5 datasets into one. In particular, only countries that were present in all 5 datasets are present in the cleaned dataset, bringing the total number of nations included in this study to 134. Further, the cleaned analysis dataset was filtered to only contain the following information from the 5 datasets: a country's name, its incarceration rate (per 100k population), its poverty rate, its unemployment rate, its violent crime rate, and its educational ranking category. Note that a nation's violent crime rate and educational ranking category were not explicit data that were provided as part of the 5 raw datasets. Instead, these were manually constructed and appended to the final dataset during the cleaning process.

In this study, a nation's violent crime rate is defined to be the sum of all the crime rates (per 100k population) available through the violent crime dataset in order to have one final value that represents the extent of violent crime in a nation. It is important to note, however, that various nations had missing values for some of the violent crime statistics, and as a result, these values were simply treated as 0 for the sake of the summation. As for the educational ranking category, as opposed to treating educational ranking as a continuous variable in the analysis that will follow, we instead decided to create 10 categories representing the 10 deciles of educational ranks present in the final dataset, based on the ranks of the 134 countries present, essentially splitting the countries into 10 equally sized categories by educational ranking. This was done in order to have a categorical variable whose coefficient will be more easily interpreted in the model present in @sec-model, as interpreting a 1 unit increase in educational ranking can be quite arbitrary, given that moving up or down one rank educationally may be more subjective. @tbl-cleaned-data showcases the top 5 countries by incarceration rate in the cleaned dataset.

@@

As it stands, by adhering to the figures above, there does not appear to be a strong correlation between any one of the variables and incarceration rates. In particular, poverty rates and unemployment rates do not showcase any particular trend in any direction. Violent crime on the other hand, does appear to show a somewhat positive correlation with incarceration rates. However what is interesting to note is that the 5 highlighted nations spanning the entire range of incarceration rates do not span a significant range of violent crime rates, at least relative to the overall range of violent crime rates across the entire dataset. The educational ranking category appears to have a negative relationship with incarceration rates, where a higher rank, i.e. a better education, seems to correspond with a lower average incarceration rate. However, somewhat interesting to note is the drop in mean incarceration rates for categories past the peak in the "111-125" ranking category. As such, given the fairly inconclusive information provided by the graphical and tabular figures above regarding the relationship between the chosen socioeconomic factors and a nation's incarceration rate, we employ a model in @sec-model to come to a more conclusive result.

@@

As the purpose of this paper is the investigation of various socioeconomic factors and their impact on a nation's incarceration rate as a means of better understanding the United States' mass incarceration problem, we will employ a model that will better enable us to infer the relationship between these various factors and incarceration rates. Then, using this information, I will attempt to predict what every nation's, including the United States', incarceration rate should be given its own individual data for each of the determinants of incarceration rate, and subsequently uncover whether these socioeconomic factors are a good estimator of incarceration rate, or whether issues regarding mass incarceration stem from a different facet.

@@

The particular model that this paper will utilize is a Bayesian multiple linear regression model of incarceration rate as a function of the socioeconomic variables present in our analysis. In particular, the model is as follows:

@@

We can see that, somewhat unexpectedly, increases in the poverty rate negatively correlate with incarceration rates, as do increases in unemployment, assuming all else held constant. On the contrary, as predicted, violent crime rates correlate positively with incarceration rates, all else held constant. As for a nation's educational ranking, it appears that there is no clear relationship between education and incarceration rates as nations in the highest category of education, "1-19", would see a much larger increase in predicted incarceration rates as they enter categories near the median educational ranking, before then seeing smaller increases near the lowest ranked nations.

@@

## Nations with extreme incarceration rates highlight the importance of cultural factors in determining their incarceration rates {#sec-culture}

@@

The results in @tbl-incarcerationsummary indicated a negative relationship between poverty rates and incarceration rates, suggesting that nations with a generally poorer population, relative to that particular nation's cost of living, and higher income inequality may find themselves with lower incarceration rates. Though this may seem counter-intuitive at first glance, a potential explanation could be that poorer nations are simply less capable of handling crime. In particular, poorer nations may not have the funding necessary to house criminals or hire a sufficient number of individuals to police the nation. For instance Kenya, who feature a poverty rate of 36.1% - well above the median - but an incarceration rate of 107 - fairly below the median, also happen to have a jail occupancy level of 173.2% as of 2022 [@citeKenyaJail]. This could help explain Kenya's relatively low incarceration rate as being the result of an inability to house more inmates.

@@

The possible explanations for this are twofold: Firstly, as was previously examined in @sec-culture, incarceration rates can be heavily dependent on cultural aspects that were not taken into account via the model. As such, certain nations with exceedingly low or exceedingly high incarceration rates may be disproportionately affecting the mean incarceration rates of their respective educational ranking categories, in turn leading to inaccurate conclusions regarding the impact of education on incarceration rates. To illustrate this point further, El Salvador, who feature the highest incarceration rate of 1086, also happen to be in the educational ranking category of "111-125", which may explain why the model predicts that this particular category is likely to raise a nation's incarceration rate by the greatest margin.

The second possible explanation could be that nations with poorer levels of education are likely to also be poorer economically, which, as discussed in @sec-povertycorruption, could lead to lower incarceration rates, in turn providing a possible explanation for why nations with the lowest levels of education also happen to feature lower incarceration rates than some nations with better education.

@@

By incorporating all insights derived from @sec-culture, @sec-povertycorruption, and @sec-educationincarceration, we arrive to the following conclusions for how the United States can best address their issue of mass incarceration:

@@

There are a number of limitations to be addressed with regards to the analysis conducted and subsequent conclusions drawn. A significant limitation pertaining to this study is the lack of up-to-date data across all nations. As was discuss is @sec-data, all raw datasets contained information on nations that may not be accurate as of 2024. In particular, many nations had data that was collected years, and in some cases decades, ago, in turn potentially rendering much of the analysis inaccurate as significant changes could have risen over the years. Moreover, not all datasets included information on all nations, and in many cases, countries that were included in some datasets featured missing data points for various columns of data. As a result, the subset of countries that were included in this study may not accurately reflect the overarching global trends as only 134 out of the 193 countries recognized by the UN [@citeUN] were taken into account during the analysis. Thus, the subsequent conclusions drawn may be biased as the remaining countries left in the study could be skewed towards more developed nations as these may be more likely to have data readily available. Moreover, due to the missing data points for countries that were included in the study, values such as the violent crime rate statistic that was calculated for each nation is inaccurate as it only included whichever information was available for that particular nation. As a result, nations which may have reported more crime rates than others would in turn bias the results to inaccurately portray such nations as having relatively more crime than they should when compared with nations that had missing data points for certain crime statistics.

A potentially significant limitation also exists with regards to the model used. In particular, the model does not take into account any interaction terms between the various independent variables when it likely should have, given the complex and intricate dynamics between the many variables at play. Though this was done to simplify the complexity of the model to allow for easier interpretations of the results, the caveat was an overall less realistic, and possibly less accurate, model. Another limitation with the model lies in its omission of cultural variations and associated factors across nations. By not grouping countries with similar cultural values or political and judicial policies, the model fails to account for factors that could greatly influence the derived results. As we have already explored in @sec-culture, varying cultural and societal norms with regards to how nations deal with crime can significantly impact their incarceration rates - a fact not taken into account by the model.

@@

Future analysis in the context of this study can be better improved by incorporating more up-to-date data. In particular, web-scraping and compiling the raw data manually as opposed to using the raw datasets used in this study could net an overall more recent, and complete, set of data points. Moreover, incorporating time series data spanning multiple years across all nations would allow us to account for time and country fixed-effects, in turn providing more conclusive data that takes into account the various cultural and societal disparities across nations and years. Another improvement that could be made is a better analysis of the model itself in order to fine-tune aspects such as the locations of the various priors, as well as exploring the possibility of adding additional factors and interaction terms to more accurately model incarceration rates.

@@

# Model details {#sec-model-details}

@@

We employ a posterior predictive check in @fig-ppcheck to evaluate how well the model fits the data. By examining the figure, it appears as though the model does not accurately fit the data, and thus amendments likely should be made as a result [@citeAlexander].

```{r}
#| label: fig-ppcheck
#| fig-cap: Examining how the model fits the data through a posterior predictive check
#| echo: false
#| fig-pos: H
#| message: false
#| warning: false

@@

```{r}
#| label: fig-postprior
#| fig-cap: Examining how the model is affected by the data through a posterior-prior comparison
#| echo: false
#| warning: false
#| fig-pos: H
#| message: false
